Epigraphy and onomastics in the Hesperia databank

### 2015-12-17 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
updated paper 22, updated paper 24
paper 24 doc to tex
added email in paper 22

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Noemí Moncunill Martí}
\date{2015-11-15}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
1.\ \ Epigraphy and onomastics in the Hesperia databank\newline
\newline

Noemí Moncunill1, Javier Velaza2

Université de Paris-Sorbonne (Paris IV)1, University of Barcelona2

Email(s): velaza@ub.edu

nmoncunill@gmail.com

velaza@ub.edu

\bigskip
\end{minipage}
\end{figure}
The first part of this work provides a general overview of the features and advantages of the digital epigraphic corpora
on the basis of the experience gained in the last years within the Hesperia project. The second part of the paper
provides a detailed presentation of the new sections available in the Hesperia databank devoted to indigenous personal
names and divinity names.

\bigskip

Keywords

Hispania, epigraphy, onomastics, digital corpora, Palaeohispanic languages and writings

\bigskip

\bigskip

\bigskip

1. Introduction.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}

The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.

The first thing that necessarily needs to be highlighted is that traditional epigraphic corpora have certain limits that
are imposed by their morphology and format. If summarizing, there are three main and most evident limitations:

a) One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
displayed one after another, with no wider possibility to link one to another than through some indices, which by the
way are rarely detailed enough. Not even in the most extended indices, as those in CIL, the connexions between
inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
connexions can be set by the users in their mind, but when the corpus displays thousands or tens of thousands of
inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.

b) Ephemeral usefulness. It looks also obvious that a traditional corpus will necessarily be always ephemeral for many
reasons: new inscriptions cannot be added, mistakes cannot be corrected, text editions cannot be improved and
modifications based on new evidence or knowledge cannot be introduced, among other limitations. Of course some of these
problems have hardly been solved by publishing some supplements, but it is widely known that these are partial
solutions only.

c) Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
they interfere in the utility of other views, to the extend of depreciating them too much. When it comes to an
epigraphic corpus that is written in a wide-known language, such as Latin, this auctoritas becomes particularly
powerful regarding the reading of the epigraphic text, its checking being not possible all the time for users,
particularly for those who are not experts on the subject. But when it comes to fragmentary languages, with very low or
even no language deciphering at all, and with many doubts on decoding, the truth is that the influence of the editor is
extremely high and involves some risks that they do not always come to terms with successfully.

In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
puts to our disposal. Although this is no longer anything of new, any attempt to use them so far has brought to an
output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
get over one-dimensional formats, an open, adaptable digital structure must be created that shall take the most out of
every single datum in the database. Besides, getting over an ephemeral usefulness requires the possibility not just to
keep adding new data, but also to connect the database with any other tool with a similar structure and to extend the
platform with any new or advisable features. Finally, the authoritative temptation must be also avoided, particularly
through systems that shall bring to users the possibility to take their own decisions in certain issues of an open
debate and, at the end of the day, to customize their own corpus.

@@

2. The Hesperia project

All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
General presentations of the project can be found in Luján - Orduña (forthcoming), Orduña - Luján (forthcoming) and
Velaza 2014.}. First of all, we need to recall these are very different materials as for their quantity, quality and
reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
Concerning this subject, see Velaza 2009.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.

 \includegraphics[width=10.987cm,height=7.835cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}

Fig. 1: Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.

@@

Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
Monumenta linguarum Hispanicarum, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
interdisciplinary and interuniversity team with researchers from the Universidad Complutense de Madrid, the Universidad
del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
for searching (http://hesperia.ucm.es/).

 \includegraphics[width=10.896cm,height=7.669cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}

Fig. 2. Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank

i) Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
corresponding to {\textquotedbl}general features{\textquotedbl}, {\textquotedbl}text{\textquotedbl},
{\textquotedbl}epigraphy and paleography{\textquotedbl}, {\textquotedbl}pictures{\textquotedbl},
{\textquotedbl}arqueological context{\textquotedbl} and {\textquotedbl}bibliography{\textquotedbl}. The databases are
linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
Catalonia), which is scheduled to be open for search by the end of 2015.

ii) A numismatic database with its specific four-tab structure: {\textquotedbl}general features{\textquotedbl},
{\textquotedbl}inscription{\textquotedbl}, {\textquotedbl}language and writing{\textquotedbl} and
{\textquotedbl}bibliography{\textquotedbl}, where all Paleohispanic mints are gathered, no matter their scripts or
languages.

iii) An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
from the identified onomastic formulae.

@@

3. The new sections on onomastics.

@@

 \includegraphics[width=11.749cm,height=6.712cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}

Fig. 3. Personal-names and divinity-names record in Hesperia databank.

@@

 \includegraphics[width=11.883cm,height=5.66cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}

Fig. 4. Personal-names and divinity-names record in Hesperia databank.

@@

 \includegraphics[width=7.031cm,height=5.359cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}

Fig. 5: Dispersal area of indigenous personal names.

 \includegraphics[width=6.98cm,height=5.262cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}

Fig. 6: Dispersal area of indigenous divinity names

 \includegraphics[width=6.32cm,height=6.045cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}

Fig. 7: Disperal areas of indigenous divinity names and personal names

@@

 \includegraphics[width=6.491cm,height=4.833cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}

Fig. 8: Dispersal area of the anthroponymic element biur.

 \includegraphics[width=6.092cm,height=4.583cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}

Fig. 9: Dispersal area of the name Tancinus.

@@

 \includegraphics[width=10.862cm,height=6.325cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}

Fig. 10: record of a possible Greek name attested in an Iberian inscription.

@@

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Epigraphy and onomastics in the Hesperia databank}
\author[barc]{Noemí Moncunill Martí \corref{first}}
\author[sorb]{Javier Velaza}
\address[sorb]{Université de Paris-Sorbonne (Paris IV)}
\address[barc]{University of Barcelona2}
\cortext[first]{Corresponding author. Email: nmoncunill@gmail.com}
\date{2015-11-15}
\begin{document}

\maketitle
\begin{abstract}
The first part of this work provides a general overview of the features and advantages of the digital epigraphic corpora
on the basis of the experience gained in the last years within the Hesperia project. The second part of the paper
provides a detailed presentation of the new sections available in the Hesperia databank devoted to indigenous personal
names and divinity names.
\end{abstract}
\keywords{Hispania, epigraphy, onomastics, digital corpora, Palaeohispanic languages and writings}

\section{Introduction}

The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}

The first thing that necessarily needs to be highlighted is that traditional epigraphic corpora have certain limits that
are imposed by their morphology and format. If summarizing, there are three main and most evident limitations: 
\begin{description}
\item[a] One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
displayed one after another, with no wider possibility to link one to another than through some indices, which by the
way are rarely detailed enough. Not even in the most extended indices, as those in CIL, the connexions between
inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
connexions can be set by the users in their mind, but when the corpus displays thousands or tens of thousands of
inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.

\item[b] Ephemeral usefulness. It looks also obvious that a traditional corpus will necessarily be always ephemeral for many
reasons: new inscriptions cannot be added, mistakes cannot be corrected, text editions cannot be improved and
modifications based on new evidence or knowledge cannot be introduced, among other limitations. Of course some of these
problems have hardly been solved by publishing some supplements, but it is widely known that these are partial
solutions only.

\item[c] Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
they interfere in the utility of other views, to the extend of depreciating them too much. When it comes to an
epigraphic corpus that is written in a wide-known language, such as Latin, this auctoritas becomes particularly
powerful regarding the reading of the epigraphic text, its checking being not possible all the time for users,
particularly for those who are not experts on the subject. But when it comes to fragmentary languages, with very low or
even no language deciphering at all, and with many doubts on decoding, the truth is that the influence of the editor is
extremely high and involves some risks that they do not always come to terms with successfully.

\end{description}
In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
puts to our disposal. Although this is no longer anything of new, any attempt to use them so far has brought to an
output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
get over one-dimensional formats, an open, adaptable digital structure must be created that shall take the most out of
every single datum in the database. Besides, getting over an ephemeral usefulness requires the possibility not just to
keep adding new data, but also to connect the database with any other tool with a similar structure and to extend the
platform with any new or advisable features. Finally, the authoritative temptation must be also avoided, particularly
through systems that shall bring to users the possibility to take their own decisions in certain issues of an open
debate and, at the end of the day, to customize their own corpus.

@@

\section{The Hesperia project}

All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
General presentations of the project can be found in \citet{orduna_implementing_????}, \citet{orduna_philology_????} and
\citet{velaza_hesperia:_2014}.}. First of all, we need to recall these are very different materials as for their quantity, quality and
reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
Concerning this subject, see \citet{velaza_epigrafiy_2009}.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}
\caption{ Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.}
\label{fig:1}
\end{figure}

@@

Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
Monumenta linguarum Hispanicarum, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
interdisciplinary and interuniversity team with researchers from the Universidad Complutense de Madrid, the Universidad
del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
for searching (\url{http://hesperia.ucm.es/}).

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}
\caption{Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank}
\label{fig:2}
\end{figure}

\begin{description}
\item[i] Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
corresponding to ``general features'', ``text'',
``epigraphy and paleography'', ``pictures'',
``arqueological context'' and ``bibliography''. The databases are
linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
Catalonia), which is scheduled to be open for search by the end of 2015.

\item[ii] A numismatic database with its specific four-tab structure: ``general features'',
``inscription'', ``language and writing'' and
``bibliography'', where all Paleohispanic mints are gathered, no matter their scripts or
languages.

\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
from the identified onomastic formulae.

\end{description}

@@

\section{The new sections on onomastics}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:3}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}
\caption{Personal-names and divinity-names record in Hesperia databank.}
\label{fig:4}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}
\caption{Dispersal area of indigenous personal names.}
\label{fig:5}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}
\caption{Dispersal area of indigenous divinity names}
\label{fig:}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}
\caption{Disperal areas of indigenous divinity names and personal names}
\label{fig:7}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
\caption{Dispersal area of the anthroponymic element biur}
\label{fig:}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}
\caption{Dispersal area of the name Tancinuse}
\label{fig:9}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
\caption{Fig. 10: record of a possible Greek name attested in an Iberian inscription.}
\label{fig:10}
\end{figure}

@@

CTS = Canonical Text Services (protocol to cite digitally-available texts in a canonical way).

EAGLE = The Europeana network of Ancient Greek and Latin Epigraphy:  \url{http://www.eagle-network.eu/}

IDEs = Integrating Digital Epigraphies project.

LGPN = The Lexicon of Greek Personal Names: \url{http://www.lgpn.ox.ac.uk/}

Pleiades = A community-built gazetteer and graph of ancient places: \url{http://pleiades.stoa.org/}

SNAP = Standards for Networking Ancient Prosopographies project: \url{http://snapdrgn.net/}

Trismegistos = Interdisciplinary portal of papyrological and epigraphical resources formerly Egypt and the Nile valley
(800 BC-AD 800): \url{http://www.trismegistos.org/}

\nocite{moncunill_ochocientos_????}
\nocite{untermann_monumenta_1975}
\nocite{vallejo_antroponimia_????}
